---
title: "Loading the shroom package"
author: "Nathan Brouwer"
date: "November 12, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Loading devtools

If you haven't already, downloaded the **devtools** package (This isn't necessary if you have downloaded it recently )

```{r, eval = FALSE}
#Only do this if you have -never- downloaded 
#devtools before
install.packages("devtools")
```

Load **devtools** into your current R session using library()
```{r, eval = F}
library(devtools)
```


## Loading shroom

If you haven't done so recently, download the **shroom** package from github.  (The function is install_github() but what it really is doing is copying the files from GitHub to your computer).
```{r, eval = FALSE}
devtools::install_github("brouwern/shroom")
```


Load **shroom** into your current R session using library()
```{r, eval = FALSE}
library(shroom)
```


## Looking at shroom data

You can check to see if it worked by loading a dataset

```{r}
data("wingscores")
```

and looking at it with summary().  Here, we'll just look at a single column that has the scores.
```{r}
summary(wingscores$wing.score)
```

and plot it

```{r}
hist(wingscores$wing.score)
```

