---
title: "Summarizing data with dplyr"
author: "Nathan Brouwer"
date: "November 12, 2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This tutorial will demonstrate how to use dplyr's  summarize() commands to calculate summary statistics.  First we'll process a dataset of female fruit flies to determine summary stats on their wing phenotype (mean, sd etc.).  We'll use the group_by() command to split the data up by the experiment groups the flies are in.  Then we'll look at a larger dataset with both male and female flies in two groups and extend the use of the group_by() function to split larger datasets up with multiple grouping variables: sex (male vs. female) and experimental group (control vs. experimental.


### Load libraries

Any libraries not already downloaded can be loaded using the code "install.packages(...)", such as "install.packages('plotrix)".

```{r}
library(shroom)  #data
library(dplyr)   #summarize() and groub_by()
library(plotrix) #function for SE
```


### Load data

The wingscores_s1F dataset has data for the 24 female flies examined by the 1st student in the dataset ("_s1F"). The "wingscores_s1" dataset has all of the 1st students data.

```{r}
data("wingscores_s1F")
data("wingscores_s1")
```

## Numeric summaries with summarize()

We can calculate the mean of a column using summarize with the mean() command within it

```{r}
wingscores_s1F %>% 
  summarize(mean = mean(wing.score))
```


summarize() is most powerful when you want multiple summary statistics worked up for the same set of data.  We can ask for both the mean and the standard deviation like this:
```{r}
wingscores_s1F %>% 
  summarize(mean = mean(wing.score),
            SD = sd(wing.score))
```

Note that the format is "summary.name = function(data.column)".



A handy function is n() which counts up the sample size.  Note that nothing goes in n()

```{r}
wingscores_s1F %>% 
  summarize(mean = mean(wing.score),
            SD = sd(wing.score),
            n = n())
```


## Numeric summaries by group

group_by() splits the dataframe up by a factor variable.  Here, I split by the 2 experimental conditions: E = "experimental" flies and C = "control" flies.

```{r}
wingscores_s1F %>%                   # data
  group_by(E.or.C) %>%               # group_by()
  summarize(mean = mean(wing.score), # summarize()
            n = n(),
            sd = sd(wing.score))
```


## Cool summarize() tricks

Here are some demonstrations of the functionality of summarize().

### Using summary functions from other packages

I can also use functions from other packages.  The plotrix package has a standard error function.

I can use this std.error() function within summarize()
```{r}
wingscores_s1F %>% 
  group_by(E.or.C) %>% 
  summarize(mean = mean(wing.score),
            n = n(),
            sd = sd(wing.score),
            se = plotrix::std.error(wing.score))
```


### Doing calculations within summarize()

I can even do math within summarize. Here, I calculate an approximate 95% confidence interval using std.error
```{r}
wingscores_s1F %>% 
  group_by(E.or.C) %>% 
  summarize(mean = mean(wing.score),
            n = n(),
            sd = sd(wing.score),
            se = plotrix::std.error(wing.score),
            CI95 = 1.96*plotrix::std.error(wing.score))
```


### Saving summarize() output to an R object.

I can save this summary table to an object if I want

```{r}
s1_females_summary <- wingscores_s1F %>% 
  group_by(E.or.C) %>% 
  summarize(mean = mean(wing.score),
            n = n(),
            sd = sd(wing.score),
            se = plotrix::std.error(wing.score),
            CI95 = 1.96*plotrix::std.error(wing.score))
```


And call it up later
```{r}
s1_females_summary
```

I can even save this to a .csv file

```{r, eval = F}
write.csv(s1_females_summary, file = "s1_summary.csv")
```


## Numeric summaries by multiple groups

You can give group_by() multiple categories.  

The "wingscores_s1" data has both males and females from both experimental conditions.

```{r}
#names of columns to examine
columns. <- c("E.or.C", "sex","wing.score")

summary(wingscores_s1[,columns.])
```


Calculate the mean for both sexes and for both experimental conditions.
```{r}
wingscores_s1 %>% 
  group_by(E.or.C,
           sex) %>% 
  summarize(mean = mean(wing.score))
```



            
